---
title: Interaction of age and gender on post-discharge quality-of-life in adult trauma
  patients in urban India – a cohort study
documentclass: article
bibliography: bibliography.bib 
output:
  pdf_document: default
  word_document: default
csl: bmcemerg.csl
---


Introduction
============
Trauma contributes to one-tenth of the global disability-adjusted life-years (DALYs), with low-and middle-income countries (LMICs) bearing a disproportionate burden of the morbidity  [@GBD2019DemographicsCollaborators2020; @Haagsma2019]. To address this burden it is important to understand the long-term outcomes of trauma and the different factors associated with these outcomes, especially in LMICs  [@Kruithof2017; @Wisborg2017; @Rios-Diaz2017]. This encompasses a range of socioeconomic outcomes including  health-related quality of life (QOL) [@Ahmed2017; @Nguyen2017; @Hossain2020; @Jagnoor2019]. 

Age and gender are associated with post-discharge QOL among trauma patients. Elderly populations and women tend to have limited access to resources, reduced social capital, disparities in support, poor health-seeking behavior, and restricted education and employment opportunities [@Amurwon2019;@Levasseur2015; @Brinda2016; @Gupta2019; @Kennedy2020]. This can shape their post-discharge well-being and outcomes after trauma [@Waqas2017; @Awang2018; @Quadir2019; @Fabricius2020]. Consequently, older age and being a women can make trauma patients more vulnerable to poorer post-discharge QOL [@Gopinath2017;@Brown2017c; @Mollayeva2018;@Rissanen2020]. There is some evidence that older women may have higher morbidity in LMIC settings [@Carmel2019; @Agrawal2014]. There is little research on the interaction between age and gender on QOL among post-discharge trauma patients in this setting. 

Understanding the interaction between age and gender on QOL may provide insights for improving trauma management and developing support services in LMIC settings [@Meara2015; @Babhulkar2019]. The aim of this study is to assess the interaction of age and gender with post-discharge QOL among adult trauma patients using the context of urban India. 

Methods
=======

Study Design
------------

This study is a prospective cohort study of trauma patients discharged from four tertiary-care hospital in urban India between November 2019 and October 2021.

Setting
-------
India accounts for nearly 20% of global trauma burden [@GBD2019DemographicsCollaborators2020]. More than one-tenth of all the DALYs in India are due to trauma and it is among the top five causes of morbidity [@Menon2019]. The patients were enrolled from the on-going Trauma Audit Filters Trial (TAFT) in four participating tertiary-care hospitals in Indian cities. [@ClinicalTrials.gov2017]. These were the Grant Medical College and Sir Jamshedjee Jeejeebhoy Hospital in Mumbai, Lok Nayak Hospital of Maulana Azad Medical College (MAMC) in Delhi, the Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM) in Kolkata and St. John's Medical College, Bengaluru. The first three are public hospitals that have nominal fees catering to patients from lower socioeconomic sections of the population, while the fourth is a charitable private hospital catering to a mix of different socioeconomic sections of the population.

Participants
------------

We include patients aged 18, which is the legal age for consent in in India [@IndianCouncilofMedicalResearch2006],  presenting to the casualty department with a history of trauma---as per the V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) [@WordHealthOrganization2016]--who are admitted and discharged alive.

Variables
---------
Age and gender were the main  variables for this study. We also included vital sign measures such as systolic blood pressure (SBP), respiratory rate (RR), heart rate (HR), oxygen saturation (SPO), Injury severity scores (ISS) and Glasgow Coma Scale (GCS) and injury etiology measures like mode of transport to hospital, type of injury, mechanism of injury, and length of hospital stay.

Outcomes
--------
Health-related quality-of-life outcome was measured using the EQ-5D Tool [@EuroQolGroup]. EQ-5D is a standardized measure of quality of life using five dimensions: mobility, self-care, usual activities, pain/discomfort, and anxiety/depression. Each dimension in the tool has three levels: no problems, some problems, and extreme problems. Additionally there is a visual analog scale (VAS) The patients were followed up at 3-months. EQ-5D tool can be administered over the telephone and has translations available in multiple Indian languages  [@EuroQolGroup]

Table 1: Description of study variables and outcomes

+----------------------+----------------------------------------------+
| Name                 | Description                                  |
+----------------------+----------------------------------------------+
| Age                  | Patient's age rounded up to closest whole    |
|                      | year                                         |
+----------------------+----------------------------------------------+
| Sex                  | Patient's sex:                               |
|                      |                                              |
|                      | Female                                       |
|                      |                                              |
|                      | Male                                         |
+----------------------+----------------------------------------------+
| Vital Signs          | Systolic blood pressure (SBP), respiratory   |
|                      | rate (RR), heart rate (HR), oxygen           |
|                      | saturation (SPO), and Glasgow Coma Scale     |
|                      | (GCS)                                        |
+----------------------+----------------------------------------------+
| Injury Etiology      | Mode of transport, type of injury, mechanism |
|                      | of injury                                    |
+----------------------+----------------------------------------------+
| Injury Severity      | Injury Severity Score                        |
+----------------------+----------------------------------------------+
| Outcomes             |                                              |
+----------------------+----------------------------------------------+
| Quality of Life      | The participant's reply to the EQ-5D         |
|                      | questionnaire and VAS on health state at 3   |
|                      | months after arrival at the study site.      |
|                      |                                              |
+----------------------+----------------------------------------------+

Data Source
-----------

Data was collected by one dedicated independent project officer in each of the hospitals who prospectively gather data on a standardized intake form for eight hours per day, five days a week, by directly observing the staff delivering trauma care. Vital signs such as systolic blood pressure, heart-rate, and oxygen saturation was recorded by the project officer independently.  The project officer was rotated daily through each eight-hour shift in the morning, evening and night. Data for the variables was collected from patient records, or from the patient or patient representatives when they are at the hospital. Additionally, the project officer followed the patient or the patient relatives at 3-months after discharge by telephone the for information on the outcomes and any missed variables.

Bias
----

There could be bias in collection of data and recording of vital signs used to calculate injury severity. Adequate training of the project officers, periodic quality control of the data with external project
officers and weekly online review meetings was done to reduce this bias.

Statistical Methods
-------------------

Multivariable linear models was used to study the association of age and gender with the QOL estimating 95% confidence intervals and denote associations with a p-value of less than 0.05 as statistically significant  [@Carey2013]. The statistical software R was used for all statistical analyses [@RCoreTeam2015].

Study Size
----------

Simulation studies indicate that as many as 25 events, observations with the outcome, and non-events or more per free parameter are required to obtain stable estimates for logistic regression [@Courvoisier2011]. Given that our logistic regression model included around 18 free parameters, we need to include 375 events and at least as many non-events.
<!-- So event is an observation with the outcome. Which outcome are you using as the basis of this calculation?-->
<!-- SD: Oh, first, I was using the variables (age, gender, etc.). If I am using the outcomes EQ-5D VAS score (low, medium, high based on cut-off scores on the general population?) and the 5 domains with 3 levels each, so its 3 + 15 so 18. 375 events plus 375 non-events. That makes it 750? -->.

Ethics
------

Ethical clearance for the data collection was obtained from the four participating hospitals as amendments to the existing ethical clearance for the on-going TAFT project (Grant Medical College & Sir J.J., Group
of Hospitals, Mumbai---No. IEC/Pharm/CT/111/A/2017, Dated 22nd August 2017; Institute of Post-Graduate Medical Education, Kolkata---Memo No. IPGME&R/IEC/2017/396, Dated 21st August 2017; Maulana Azad Medical
College, New Delhi-F.1/IEC/MAMC/(53/2/2016/No.97), Dated 3rd August 2016; St. John's Medical College, Bengaluru---No. IEC/1/671/2017, Dated 24th August, 2017). Waiver of informed consent was granted for collection of clinical data which was routinely collected for the patients, as they were all admitted after trauma, often arriving in an altered level of consciousness and in severe physical and psychological distress. The amendment granted permission to collect the additional data necessary for this study (Grant  Medical College & Sir J.J., Group of Hospitals, Mumbai---No. IEC/Pharm/CT/2059/2019, Dated 16th September 2019; St. John's Medical College, Bengaluru---No. IEC/1/530/2019, Dated 25th June, 2019).

Data management
---------------

Each center was assigned a center identification number and each patient a locally unique study identification number. Project officers first entered data on paper without any personal identification data. The project officers then transferred this data to an electronic format using a dedicated data entry application. The electronic data did not include any direct identifiers such as name, hospital record number, and telephone numbers. The only way to link an electronic record to a paper intake form was by combining the record's hospital and study identification numbers. Paper forms were kept locally at each center for the duration required by locally applicable laws and regulations, or at least five years, whichever is longest. The adequacy of their storage was the responsibility of the principal investigator at each center. Care was taken taken to ensure that at no time where they stored with less than reasonable care.

Results
========
```{r dataset, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}

## Required packages
library(assertthat)
library(epiDisplay)
library(dplyr)
library(knitr)
library(qwraps2)
library(tidyverse)
library(tableone)
library(bengaltiger)
library(graphics)
library(caret)
library(emmeans)
library(car)
library(effects)
library(ordinal)
library(scales)


##Source functions
source("R/import_data.R")
source("R/clean_data.R")
source("R/describe_missing_data.R")
source("R/describe_data.R")
source("R/finalize_dataset.R")
source("R/create_sample_characteristics_table.R")

## Import dataset (Using import_data function)
dirty.data <- import_data("taft-dataset-collated-20200608232208.csv") 

## Clean data (Using clean_data function)
clean.data <- clean_data(dirty.data)

## Data description
data.description <- describe_data(clean.data)

## Creating dataset with variables selected for this study
final.dataset <- finalize_dataset(clean.data)

## Describing missing data (using describe_missing_data)
descriptions <- describe_missing_data(final.dataset)


##Total patients
total.patients <- nrow(clean.data)

##Missing data
missing.patients <- descriptions[["total.incomplete"]][["n"]]

##Final Data
final.patients <- total.patients - missing.patients

```

```{r datasummarytable, include = TRUE, echo=FALSE, message=FALSE, warning=FALSE}
## Create Sample Characteristics Table
sample.charac <- create_sample_characteristics_table(final.dataset)

## Print table in manuscript
kable(sample.charac$table, caption = sample.charac$caption, format = "pandoc")

```

```{r interaction, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
## EQ-5D Index Score

##Complete patients
hs.complete <- NROW(final.dataset$eq5dhs) - descriptions[["eq5dhs"]][["n"]]

##Mean HS score
hs.mean <- mean(final.dataset$eq5dhs, na.rm = T)

##Proportion with No problems in domains
final.dataset$eq5dm

BCI_percent <- BCI / rowSums(BCI) * 100




##Creating a model with age and gender
model1 <- lm(eq5dhs ~ age*sex, data = final.dataset)
#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term

summary(model1)
##Interaction not significant 

##Calculating the VIF using the car package
vif(model1)

##It is above 5 for age (7.5) and the interaction factor (8.8)

##Standardizing age and interaction factor by Centering: subtracting by mean and dividing by SD
final.dataset$centred.age <- (final.dataset$age - mean(final.dataset$age, na.rm = TRUE)) / sd(final.dataset$age, na.rm = TRUE)

##New model using centred age
cent.model1 <- lm(eq5dhs ~ centred.age*sex, data = final.dataset)

summary(cent.model1)
##Estimate values do change a bit 

vif(cent.model1)
##But all the VIFs are less than 5

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)

model2 <- lm(eq5dhs ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + centred.age*sex, data = final.dataset)

summary(model2)
##Interaction not significant. Interaction coefficient is -1.13 and after adjusting it is -1.59.

vif(model2)
##All values below 5

##Checking for individual slopes by sex  using the emtrends function

emtrends(model2, ~ sex, var = "centred.age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(model2, pairwise ~ sex, var = "centred.age")

##Interaction of age and gender is not significant. 

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(centred.age = seq(-1.32,3.10, by = 0.1), sex = c("Female","Male")))
plot1 <- emmip(model2, sex ~ centred.age,at = categories1, CIs=TRUE)

##saving plot
jpeg("plot-1.Health-Status.jpeg", quality = 100)
emmip(model2, sex ~ centred.age,at = categories1, CIs=TRUE)
dev.off()

##Overall, with older age it is men who tend to have low QOL, while young women have lower QOL than young men. However, the large overlap of confidence interval indicates EQ5D Health status across age is not statistically different for men and women after adjusting. 

##Dividing age into categories and comparing interaction over sex

##Making age into categorical
final.dataset$agegrp <- cut(final.dataset$age, c(17,32,49,89), labels = c("18-32","33-49","50 and above"))

##Making male the reference group and age group 18-32 the reference group as they had the highest QOL

final.dataset$agegrp2 <- relevel(final.dataset$agegrp, ref = "18-32")
final.dataset$sex2 <- relevel(final.dataset$sex, ref = "Male")

model3 <- lm(eq5dhs ~ sex2*agegrp2, data = final.dataset)

summary(model3)

## Younger men aged (aged 18-25) had significantly higher QOL scores than the middle aged men (aged 33-49). Older women and younger women had better QOL scores than younger women. All other groups had lower scores than younger men.

vif(model3)
##All values are less than 5

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
model4 <- lm(eq5dhs ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset)

summary(model4)

##As before above, younger men aged (aged 18-25) had significantly higher QOL scores than the middle aged men (aged 33-49) (Estimate -4.86). Older women (Estimate 3.5) had much better QOL scores than younger men, while younger women had marginally better QOL scores than younger men (Estimate 0.5). All other groups had lower scores than younger men. 

vif(model4)
##All values are less than 5

##Calculating the values of QOL by all possible combinations of the age and sex categories
cat.values <- emmeans(model4, ~ sex2*agegrp2)

##Estimating the differences in interactions across age and sex categories interaction using contrast function and argument "revpairwise" as Female is the default in the data set

contrast(cat.values, "revpairwise", by ="agegrp2")

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
plot2 <- emmip(model4, sex2 ~ agegrp2, CIs = TRUE)

##Saving plot
jpeg("plot-2.Health-Status.jpeg", quality = 100)
emmip(model4, sex2 ~ agegrp2, CIs = TRUE)
dev.off()

##Middle aged women (33-49) seem to have the lowest QOL Scores.

##EQ5D Domains
## Mobility

##Creating a model with age and gender using the polr function
m.model1 <- polr(eq5dm ~ age*sex, data = final.dataset)

#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term
summary(m.model1)
m.model.ci <- confint.default(m.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(m.model1), m.model.ci))

##For males the odds of having some mobility problems was 1.05 times that of females. For every one year increase in age the odds of some mobility problems increases by 1.019 times (i.e. increases by 1.9%).The interaction factor reduces the odds of having some mobility problems by 1% [(1- 0.99*100)].??

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
m.model2 <- polr(eq5dm ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset)


summary(m.model2)
m.model2.ci <- confint.default(m.model2)



##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(m.model2), m.model2.ci))

##After adjusting, for males the odds of having some mobility problems was 1.14 times that of females. For every one year increase in age the odds of some mobility problems increases by 1.012 times (i.e. increases by 1.2%).The interaction factor reduces the odds of having some mobility problems by 1% [(1- 0.99*100)].??



##Checking for individual slopes by sex  using the emtrends function
emtrends(m.model2, ~ sex, var = "age")


##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(m.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.0037

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
m.plot1 <- emmip(m.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Mobility.jpeg", quality = 100)
emmip(m.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)
dev.off()

##Tried the scales package and the scale function in emmip scales = exp_trans(), ylim = c(0, 2) or scales = log10_trans(). It doesn't seem to work

##Using effects package for a stacked plot
jpeg("plot-1.Mobility_Stacked.jpeg", quality = 100)
plot(Effect(focal.predictors = c("age","sex"), mod = m.model2), 
     rug = FALSE,
     style="stacked")
dev.off()

##Using effects package for a split groups plot
jpeg("plot-1.Mobility_Split.jpeg", quality = 100)
plot(Effect(focal.predictors = c("age","sex"), mod = m.model2), 
     rug = FALSE)
dev.off()

##Dividing age into categories and comparing interaction over sex
m.model3 <- polr(eq5dm ~ sex2*agegrp2, data = final.dataset)

summary(m.model3)
m.model3.ci <- confint.default(m.model3)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(m.model3), m.model3.ci))

## For younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having some mobility problems was 1.08, 1.04, and 1.06 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) having some mobility problems was 1.51 and 1.47 times than younger males. For every one year increase in age the odds of some mobility problems increases by 1.012 times (i.e. increases by 1.2%).

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
m.model4 <- polr(eq5dm ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset)

summary(m.model4)
m.model4.ci <- confint.default(m.model4)

##As the estimates are ordered log odds converting them to odds ratios
mobility.odds <- exp(cbind(OR = coef(m.model4), m.model4.ci))

##OR
m.young.fem <- mobility.odds["sex2Female", "OR"]
m.middle.fem <- mobility.odds["sex2Female:agegrp233-49", "OR"]
m.old.fem <- mobility.odds["sex2Female:agegrp250 and above", "OR"]
m.middle.mal <- mobility.odds["agegrp233-49", "OR"]
m.old.mal <- mobility.odds["agegrp250 and above", "OR"]

## After adjusting, for younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having some mobility problems was 1.05, 1.04, and 1.04 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) having some mobility problems was 1.42 and 1.57 times than younger males. All groups had higher odds of having some mobility problems than young males.

##Calculating the values of mobility by all possible combinations of the age and sex categories
 m.cat.values <- emmeans(m.model4, ~ sex2*agegrp2, type = "response")

##Estimating the differences in interactions across age and sex categories interaction using pairs function as these are logs of odds

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
m.plot2 <- emmip(m.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)

##Saving plot
jpeg("plot-2.Mobility.jpeg", quality = 100)
emmip(m.model4, sex2 ~ agegrp2, CIs = TRUE)
dev.off()

##Using effects package for a stacked plot
jpeg("plot-2.Mobility_Stacked.jpeg", quality = 100)
plot(Effect(focal.predictors = c("agegrp2","sex2"), mod = m.model4), 
     rug = FALSE,
     style="stacked")
dev.off()

##Using effects package for a split groups plot
jpeg("plot-2.Mobility_Split.jpeg", quality = 100)
plot(Effect(focal.predictors = c("agegrp2","sex2"), mod = m.model4), 
     rug = FALSE)
dev.off()


##Self-care

##Creating a model with age and gender using the polr function
sc.model1 <- polr(eq5dsc ~ age*sex, data = final.dataset)

#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term
summary(sc.model1)
sc.model.ci <- confint.default(sc.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(sc.model1), sc.model.ci))

##For males the odds of having some self-care problems was 0.63 times that of females. For every one year increase in age the odds of some mobility problems increases by 0.5%.The interaction factor does not change the odds of having self-care [1.008].

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
sc.model2 <- polr(eq5dsc ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset)


summary(sc.model2)
sc.model2.ci <- confint.default(sc.model2)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(sc.model2), sc.model2.ci))

##After adjusting, for males the odds of having some mobility problems was 0.57 times that of females. For every one year increase in age the odds of some mobility problems increased by 1 times (i.e. increases by 0.99%).The interaction factor did not change the odds of having self-care [1.008].

##Checking for individual slopes by sex  using the emtrends function
emtrends(sc.model2, ~ sex, var = "age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(sc.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.0085

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
sc.plot1 <- emmip(sc.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Self-care.jpeg", quality = 100)
emmip(sc.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)
dev.off()

##Dividing age into categories and comparing interaction over sex
sc.model3 <- polr(eq5dsc ~ sex2*agegrp2, data = final.dataset)

summary(sc.model3)
sc.model3.ci <- confint.default(sc.model3)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(sc.model3), sc.model3.ci))

## For younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having some problems performing self-care were 1.27, 1.22, and 0.92 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) having some problems performing self-care was 1.46 and 1.43 times than younger males. 

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
sc.model4 <- polr(eq5dsc ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset)

summary(sc.model4)
sc.model4.ci <- confint.default(sc.model4)

##As the estimates are ordered log odds converting them to odds ratios
sc.odds <- exp(cbind(OR = coef(sc.model4), sc.model4.ci)) 

##OR
sc.young.fem <- sc.odds["sex2Female", "OR"]
sc.middle.fem <- sc.odds["sex2Female:agegrp233-49", "OR"]
sc.old.fem <- sc.odds["sex2Female:agegrp250 and above", "OR"]
sc.middle.mal <- sc.odds["agegrp233-49", "OR"]
sc.old.mal <- sc.odds["agegrp250 and above", "OR"]

## After adjusting, for younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having some problems performing self-care was 1.32, 1.16, and 0.78 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) some problems performing self-care was 1.33 and 1.07 times than younger males. Only older females had lower odds of problems with self-care than young males.

##Calculating the values of mobility by all possible combinations of the age and sex categories
 sc.cat.values <- emmeans(sc.model4, ~ sex2*agegrp2, type = "response")

##Estimating the differences in interactions across age and sex categories interaction using pairs function as these are logs of odds

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
sc.plot2 <- emmip(sc.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)

##Saving plot
jpeg("plot-2.Self-care.jpeg", quality = 100)
emmip(sc.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)
dev.off()

##Usual activities

##Creating a model with age and gender using the polr function
ua.model1 <- polr(eq5dua ~ age*sex, data = final.dataset)

#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term
summary(ua.model1)
ua.model.ci <- confint.default(ua.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ua.model1), ua.model.ci))

##For males the odds of having some problems in performing usual activities was 0.79 times that of females. For every one year increase in age the odds of some mobility problems increases by 1.2%.The interaction factor does not affect the odds of having self-care [1.002].

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
ua.model2 <- polr(eq5dua ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset)


summary(ua.model2)
ua.model2.ci <- confint.default(ua.model2)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ua.model2), ua.model2.ci))

##After adjusting, for males the odds of having some mobility problems was 0.73 times that of females. For every one year increase in age the odds of some mobility problems increased by 1 times (i.e. increases by 0.99%).The interaction factor did not change the odds of having self-care [1.005].

##Checking for individual slopes by sex  using the emtrends function
emtrends(ua.model2, ~ sex, var = "age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(ua.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.0053

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
ua.plot1 <- emmip(ua.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Usual-activities.jpeg", quality = 100)
emmip(ua.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)
dev.off()

##Dividing age into categories and comparing interaction over sex
ua.model3 <- polr(eq5dua ~ sex2*agegrp2, data = final.dataset)

summary(ua.model3)
ua.model3.ci <- confint.default(ua.model3)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ua.model3), ua.model3.ci))

## For younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having some problems performing usual activities were 1.24, 0.92, and 0.89 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) having some problems performing self-care was 1.68 and 1.53 times than younger males. 

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
ua.model4 <- polr(eq5dua ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset)

summary(ua.model4)
ua.model4.ci <- confint.default(ua.model4)

##As the estimates are ordered log odds converting them to odds ratios
ua.odds <- exp(cbind(OR = coef(ua.model4), ua.model4.ci)) 

##OR
ua.young.fem <- ua.odds["sex2Female", "OR"]
ua.middle.fem <- ua.odds["sex2Female:agegrp233-49", "OR"]
ua.old.fem <- ua.odds["sex2Female:agegrp250 and above", "OR"]
ua.middle.mal <- ua.odds["agegrp233-49", "OR"]
ua.old.mal <- ua.odds["agegrp250 and above", "OR"]

## After adjusting, for younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having some problems performing usual activities was 1.32, 0.79, and 0.80 times than of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) some problems performing usual activities was 1.69 and 1.24 times than young males. Only middle aged and older females had lower odds of having problems performing usual activities than young males.

##Calculating the values of mobility by all possible combinations of the age and sex categories
 ua.cat.values <- emmeans(ua.model4, ~ sex2*agegrp2, type = "response")

##Estimating the differences in interactions across age and sex categories interaction using pairs function as these are logs of odds

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
ua.plot2 <- emmip(ua.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)

##Saving plot
jpeg("plot-2.Usual-activities.jpeg", quality = 100)
emmip(ua.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)
dev.off()

##Pain and discomfort

##Creating a model with age and gender using the polr function
pd.model1 <- polr(eq5dpd ~ age*sex, data = final.dataset)

#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term
summary(pd.model1)
pd.model.ci <- confint.default(pd.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(pd.model1), pd.model.ci))

##For males the odds of having any form of pain or discomfort was 0.32 times that of females. For every one year increase in age the odds of any pain or discomfort increases by 1%.The interaction factor reduces the odds of  affect the odds of pain or discomfort by 22% [1-1.022*100].

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
pd.model2 <- polr(eq5dpd ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset)

summary(pd.model2)
pd.model2.ci <- confint.default(pd.model2)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(pd.model2), pd.model2.ci))

##After adjusting, for males the odds of having pain or discomfort was 0.32 times that of females. For every one year increase in age the odds of any pain or discomfort increased by 1%.The interaction factor reduces the odds of  affect the odds of pain or discomfort by 21% [1-1.021*100]..

##Checking for individual slopes by sex  using the emtrends function
emtrends(pd.model2, ~ sex, var = "age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(pd.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.021

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
pd.plot1 <- emmip(pd.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Pain-Discomfort.jpeg", quality = 100)
emmip(pd.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)
dev.off()

##Dividing age into categories and comparing interaction over sex
pd.model3 <- polr(eq5dpd ~ sex2*agegrp2, data = final.dataset)

summary(pd.model3)
pd.model3.ci <- confint.default(pd.model3)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(pd.model3), pd.model3.ci))

## For younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having some pain or discomfort were 1.69, 0.62, and 0.62 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) having any pain was 2.07 and 2.03 times than younger males. 

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
pd.model4 <- polr(eq5dpd ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset)

summary(pd.model4)
pd.model4.ci <- confint.default(pd.model4)

##As the estimates are ordered log odds converting them to odds ratios
pd.odds <- exp(cbind(OR = coef(pd.model4), pd.model4.ci)) 

##OR
pd.young.fem <- pd.odds["sex2Female", "OR"]
pd.middle.fem <- pd.odds["sex2Female:agegrp233-49", "OR"]
pd.old.fem <- pd.odds["sex2Female:agegrp250 and above", "OR"]
pd.middle.mal <- pd.odds["agegrp233-49", "OR"]
pd.old.mal <- pd.odds["agegrp250 and above", "OR"]

## After adjusting, for younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having any pain was 1.7, 0.58, and 0.60 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) some problems performing usual activities was 2.1 and 1.9 times than younger males. Only Middle aged and older males had lower odds of having any form of pain or discomfort than younger males.

##Calculating the values of mobility by all possible combinations of the age and sex categories
pd.cat.values <- emmeans(pd.model4, ~ sex2*agegrp2, type = "response")

##Estimating the differences in interactions across age and sex categories interaction using pairs function as these are logs of odds

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
pd.plot2 <- emmip(pd.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)

##Saving plot
jpeg("plot-2.Pain-discomfort.jpeg", quality = 100)
emmip(pd.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)
dev.off()

##Anxiety & Depression

##Creating a model with age and gender using the polr function
ad.model1 <- polr(eq5dad ~ age*sex, data = final.dataset)

#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term
summary(ad.model1)
ad.model.ci <- confint.default(ad.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ad.model1), ad.model.ci))

##For males the odds of having any form of anxiety and depression  was 0.32 times that of females. For every one year increase in age the odds of any anxiety and depression increases by 1%.The interaction factor does not seem to affect odds of anxiety or depression [1-1.0048*100].

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
ad.model2 <- polr(eq5dad ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset)

summary(ad.model2)
ad.model2.ci <- confint.default(ad.model2)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ad.model2), ad.model2.ci))

##After adjusting, for males the odds of having any anxiety or depression was 0.77 times that of females. For every one year increase in age the odds of any pain or discomfort increased by 1%.The interaction factor does not show to affect the odds of anxiety or depression [1-1.003*100]..

##Checking for individual slopes by sex  using the emtrends function
emtrends(ad.model2, ~ sex, var = "age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(ad.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.032

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
ad.plot1 <- emmip(ad.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Anxiety-Depresssion.jpeg", quality = 100)
emmip(ad.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)
dev.off()

##Dividing age into categories and comparing interaction over sex
ad.model3 <- polr(eq5dad ~ sex2*agegrp2, data = final.dataset)

summary(ad.model3)
ad.model3.ci <- confint.default(ad.model3)

##As the estimates are ordered log odds converting them to odds ratios
ad.odds <-exp(cbind(OR = coef(ad.model3), ad.model3.ci))

##OR
ad.young.fem <- ad.odds["sex2Female", "OR"]
ad.middle.fem <- ad.odds["sex2Female:agegrp233-49", "OR"]
ad.old.fem <- ad.odds["sex2Female:agegrp250 and above", "OR"]
ad.middle.mal <- ad.odds["agegrp233-49", "OR"]
ad.old.mal <- ad.odds["agegrp250 and above", "OR"]

## For younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having any anxiety and depression were 1.51, 0.47, and 1.03 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) having any anxiety and depression was 1.79 and 1.1 times than younger males. 

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
ad.model4 <- polr(eq5dad ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset)

summary(ad.model4)
ad.model4.ci <- confint.default(ad.model4)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ad.model4), ad.model4.ci)) 

## After adjusting, for younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having any anxiety and depression was 1.39, 0.48, and 1.10 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) some anxiety and depression was 1.59 and 0.84 times than younger males. Only Older males and middle aged females had lower  odds of having any form of anxiety and depression than younger males.

##Calculating the values of mobility by all possible combinations of the age and sex categories
ad.cat.values <- emmeans(ad.model4, ~ sex2*agegrp2, type = "response")

##Estimating the differences in interactions across age and sex categories interaction using pairs function as these are logs of odds

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
ad.plot2 <- emmip(ad.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)

##Saving plot
jpeg("plot-2.Anxiety-Depression.jpeg", quality = 100)
emmip(ad.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)
dev.off()

```
Out of a total of `r dataset$total.patients` trauma patients enrolled, `r dataset$missing.patients` were excluded for missing data. The final cohort of `r dataset$final.patients` included in this study had a median age of `r data.description$age.median` years (IQR: `r data.description$age.iqr.upper`-`r data.description$age.iqr.lower`) with `r data.description$proportion.male` per cent of patients were male. The most common mechanisms of injury were  `r data.description$moi1` (`r data.description$p.road` per cent). Majority of the injuries were blunt (`r data.description$proportion.blunt` per cent) and `r data.description$proportion.transferred` per cent were transferred to the study hospitals. The all cause 30-day mortality was `r  data.description$p.outcome.dead` per cent. The median severity score was x. Details of the study population is given in Table 2.


EQ-5D Scores
------------
`r interaction$hs.complete` completed the EQ-5D questionnaire at 3 months after discharge. The median index score was 80.0 [IQR: 65.0-90.0]. Three months after discharge, just over half the patients (54.8%) reported no problems with mobility while two-thirds reported no problems with self-care. While less than half the patients (43.0%) could carry on usual activities without any problems and only one-third of the patients (34.7%) reported no pain or discomfort after three-months of post-discharge. Around 40% of the patients reported experiencing some form of anxiety or depression. The overall EQ-5D scores are provided in Table 3.

EQ-5D in relation to gender and age
-------------------------------
Comparing EQ5D index scores across males and females across age groups, after adjusting for injury etiology, vitals, and severity, indicate that young males aged (aged 18-32) had significantly higher scores (p = 0.02) than the middle aged (aged 33-49) males. Only old females (aged 50 and above) had higher scores than young males All other groups had lower scores than young males (Table 4).

In the mobility dimension, keeping young males as the reference group, the adjusted odds of having mobility problems was `r interaction$m.young.fem` in young females (aged 18-32), `r interaction$m.middle.fem` middle aged females (aged 33-49), and `r interaction$m.old.fem` old females (aged 50 and above). In middle aged males (aged 33-49) the odds was `r interaction$m.middle.mal` and in old males (aged 50 and above) the odds was `r interaction$m.old.mal`. Thus, all groups had higher odds of having some mobility problems than young males.

Assessing the self-care dimension, the adjusted odds of having problems performing self-care tasks was `r interaction$sc.young.fem` in young females, `r interaction$sc.middle.fem` in middle aged females, and `r interaction$sc.old.fem` old females compared to young males. On the other hand, the odds of having these problems in middle aged males was `r interaction$sc.middle.mal` and in old males it was `r interaction$sc.old.mal`times than young males. Only old females had lower odds of reporting problems with self-care than young males.

In the dimension of being able to perform usual activities, the adjusted odds of having problems performing was `r interaction$ua.young.fem` in young females, `r interaction$ua.middle.fem` in middle aged females, and `r interaction$ua.old.fem` old females with respect to young males. The odds of having these problems was `r interaction$sc.middle.mal` in middle aged males and  `r interaction$sc.old.mal` times in in old males. Thus, middle aged and old females had lower odds of having problems performing usual activities than young males.

The adjusted odds of having any form of pain and discomfort was `r interaction$pd.young.fem` in young females, `r interaction$pd.middle.fem` in middle aged females, and `r interaction$pd.old.fem` old females compared to young males. Among middle aged and old males these odds were `r interaction$pd.middle.mal` and  `r interaction$pd.old.mal` respectively. Thus, middle aged and old females had lower odds of having problems performing usual activities than young males. So, middle aged and older males had lower odds of having any form of pain or discomfort than young males. Females across all ages had higher odds of reporting pain and discomfort. 

In the anxiety and depression dimension, keeping young males as the reference group, the adjusted odds of reporting any form of anxiety or depression was `r interaction$ad.young.fem` in young females, `r interaction$ad.middle.fem` middle aged females, and `r interaction$ad.old.fem` old females. In middle aged males the odds was `r interaction$ad.middle.mal` and in old males the odds was `r interaction$ad.old.mal`. Thus, all groups had higher odds of having some mobility problems than young males. So, Old males and middle aged females had lower odds of reporting having any form of anxiety and depression than young males.

Discussion
==========



Conclusion
==========



References
==========
