---
title: Interaction of age and gender on post-discharge quality-of-life in adult trauma
  patients in urban India – a cohort study
documentclass: article
output:
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: bmcemerg.csl
---


Introduction
============
Trauma contributes to one-tenth of the global disability-adjusted life-years (DALYs), with low-and middle-income countries (LMICs) bearing a disproportionate burden of the morbidity  [@GBD2019DemographicsCollaborators2020; @Haagsma2019]. To address this burden it is important to understand the long-term outcomes of trauma and the different factors associated with these outcomes, especially in LMICs  [@Kruithof2017; @Wisborg2017; @Rios-Diaz2017]. This encompasses a range of socioeconomic outcomes including  health-related quality of life (QOL) [@Ahmed2017; @Nguyen2017; @Hossain2020; @Jagnoor2019]. 

Age and gender are associated with post-discharge QOL among trauma patients. Elderly populations and women tend to have limited access to resources, reduced social capital, disparities in support, poor health-seeking behavior, and restricted education and employment opportunities [@Amurwon2019;@Levasseur2015; @Brinda2016; @Gupta2019; @Kennedy2020]. This can shape their post-discharge well-being and outcomes after trauma [@Waqas2017; @Awang2018; @Quadir2019; @Fabricius2020]. 

Consequently, older age and being a women can make trauma patients more vulnerable to poorer post-discharge QOL [@Gopinath2017;@Brown2017c; @Mollayeva2018;@Rissanen2020]. There is some evidence that older women may have higher morbidity in LMIC settings [@Carmel2019; @Agrawal2014], but there is little research on the interaction between age and gender on QOL among post-discharge trauma patients in this setting. 

Understanding the interaction between age and gender on QOL may provide insights for improving trauma management and developing support services in LMIC settings [@Meara2015; @Babhulkar2019]. The aim of this study is to assess the interaction of age and gender with post-discharge QOL among adult trauma patients using the context of urban India. 

Methods
=======

Study Design
------------

This study is a cross-sectional study using data from an ongoing interrupted time series trial of trauma patients discharged from four tertiary-care hospital in urban India between November 2019 and May 2021.

Setting
-------
India accounts for nearly 20% of global trauma burden [@GBD2019DemographicsCollaborators2020]. More than one-tenth of all the DALYs in India are due to trauma and it is among the top five causes of morbidity [@Menon2019]. The patients were enrolled from the on-going Trauma Audit Filters Trial (TAFT) in four participating tertiary-care hospitals in Indian cities [@ClinicalTrials.gov2017]. These were the Grant Medical College and Sir Jamshedjee Jeejeebhoy Hospital in Mumbai, Lok Nayak Hospital of Maulana Azad Medical College (MAMC) in Delhi, the Institute of Post-Graduate Medical Education and Research and Seth Sukhlal Karnani Memorial Hospital (SSKM) in Kolkata and St. John's Medical College, Bengaluru. The first three are public hospitals that have nominal fees catering to patients from lower socioeconomic sections of the population, while the fourth is a charitable private hospital catering to a mix of different socioeconomic sections of the population.

Participants
------------

We include patients aged 18, which is the legal age for consent in in India [@IndianCouncilofMedicalResearch2006],  presenting to the casualty department with a history of trauma---as per the V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) [@WordHealthOrganization2016]--who are admitted and discharged alive.

Variables
---------
Age and gender were the main  variables for this study. We also included vital sign measures such as systolic blood pressure (SBP), respiratory rate (RR), heart rate (HR), oxygen saturation (SPO), Injury severity scores (ISS) and Glasgow Coma Scale (GCS) and injury etiology measures like mode of transport to hospital, type of injury, mechanism of injury, and length of hospital stay.

Outcomes
--------
Health-related quality-of-life outcome was measured using the EQ-5D Tool [@EuroQolGroup]. EQ-5D is a standardized measure of quality of life using five dimensions: mobility, self-care, usual activities, pain/discomfort, and anxiety/depression. Each dimension in the tool has three levels: no problems, some problems, and extreme problems. Additionally there is a visual analog scale (VAS) The patients were followed up at 3-months. EQ-5D tool can be administered over the telephone and has translations available in multiple Indian languages  [@EuroQolGroup]

Table 1: Description of study variables and outcomes

+----------------------+----------------------------------------------+
| Name                 | Description                                  |
+----------------------+----------------------------------------------+
| Age                  | Patient's age rounded up to closest whole    |
|                      | year                                         |
+----------------------+----------------------------------------------+
| Sex                  | Patient's sex:                               |
|                      |                                              |
|                      | Female                                       |
|                      |                                              |
|                      | Male                                         |
+----------------------+----------------------------------------------+
| Vital Signs          | Systolic blood pressure (SBP), respiratory   |
|                      | rate (RR), heart rate (HR), oxygen           |
|                      | saturation (SPO), and Glasgow Coma Scale     |
|                      | (GCS)                                        |
+----------------------+----------------------------------------------+
| Injury Etiology      | Mode of transport, type of injury, mechanism |
|                      | of injury                                    |
+----------------------+----------------------------------------------+
| Injury Severity      | Injury Severity Score                        |
+----------------------+----------------------------------------------+
| Outcomes             |                                              |
+----------------------+----------------------------------------------+
| Quality of Life      | The participant's reply to the EQ-5D         |
|                      | questionnaire and VAS on health state at 3   |
|                      | months after arrival at the study site.      |
|                      |                                              |
+----------------------+----------------------------------------------+

Data Source
-----------

Data was collected by one dedicated independent project officer in each of the hospitals who prospectively gather data on a standardized intake form for eight hours per day, five days a week, by directly observing the staff delivering trauma care. Vital signs such as systolic blood pressure, heart-rate, and oxygen saturation was recorded by the project officer independently.  The project officer was rotated daily through each eight-hour shift in the morning, evening and night. Data for the variables was collected from patient records, or from the patient or patient representatives when they are at the hospital. Additionally, the project officer followed the patient or the patient relatives at 3-months after discharge by telephone the for information on the outcomes and any missed variables.

Bias
----

There could be bias in collection of data and recording of vital signs used to calculate injury severity. Adequate training of the project officers, periodic quality control of the data with external project officers and weekly online review meetings was done to reduce this bias.

Statistical Methods
-------------------

Multivariable linear models was used to study the association of age and gender with the EQ-5D index score and multivariable logistic regression was used to measure the association with the different domains of EQ-5D which have multilevel responses. Estimating 95% confidence intervals and denote associations with a p-value of less than 0.05 as statistically significant [@Carey2013]. The statistical software R was used for all statistical analyses [@RCoreTeam2015].

Study Size
----------

Previous studies indicate that as many as 25 events per variable is adequate to obtain stable estimates for linear regression [@Courvoisier2011; @Austin2015]. Given that our model includes around 18 free parameters, we need a minimum sample size of around 450 for the EQ-5D index score.  For multivariable logistic regression events per variable is ratio of the smallest number of observations in the multinomial outcome categories divided by the number of free parameters and should be at least 10 [@DeJong2019]. With 18 free parameters, requiring at least 180 events per category of the EQ5D domains, if there are inadequate samples, the categories would be combined as "Having no problems" and "Having any problems". 

Ethics
------

Ethical clearance for the data collection was obtained from the four participating hospitals as amendments to the existing ethical clearance for the on-going TAFT project (Grant Medical College & Sir J.J., Group of Hospitals, Mumbai---No. IEC/Pharm/CT/111/A/2017, Dated 22nd August 2017; Institute of Post-Graduate Medical Education, Kolkata---Memo No. IPGME&R/IEC/2017/396, Dated 21st August 2017; Maulana Azad Medical College, New Delhi-F.1/IEC/MAMC/(53/2/2016/No.97), Dated 3rd August 2016; St. John's Medical College, Bengaluru---No. IEC/1/671/2017, Dated 24th August, 2017). Waiver of informed consent was granted for collection of clinical data which was routinely collected for the patients, as they were all admitted after trauma, often arriving in an altered level of consciousness and in severe physical and psychological distress. The amendment granted permission to collect the additional data necessary for this study (Grant  Medical College & Sir J.J., Group of Hospitals, Mumbai---No. IEC/Pharm/CT/2059/2019, Dated 16th September 2019; St. John's Medical College, Bengaluru---No. IEC/1/530/2019, Dated 25th June, 2019).

Data management
---------------

Each center was assigned a center identification number and each patient a locally unique study identification number. Project officers first entered data on paper without any personal identification data. The project officers then transferred this data to an electronic format using a dedicated data entry application. The electronic data did not include any direct identifiers such as name, hospital record number, and telephone numbers. The only way to link an electronic record to a paper intake form was by combining the record's hospital and study identification numbers. Paper forms were kept locally at each center for the duration required by locally applicable laws and regulations, or at least five years, whichever is longest. The adequacy of their storage was the responsibility of the principal investigator at each center. Care was taken taken to ensure that at no time where they stored with less than reasonable care.

Results
========
```{r dataset, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}

## Required packages
library(assertthat)
library(epiDisplay)
library(dplyr)
library(knitr)
library(qwraps2)
library(tidyverse)
library(tableone)
library(bengaltiger)
library(graphics)
library(caret)
library(emmeans)
library(car)
library(effects)
library(ordinal)
library(scales)
library(tinytex)
library(rmarkdown)

##Source functions
source("R/import_data.R")
source("R/clean_data.R")
source("R/describe_missing_data.R")
source("R/describe_data.R")
source("R/finalize_dataset.R")
source("R/create_sample_characteristics_table.R")

## Import dataset (Using import_data function)
dirty.data <- import_data("taft-dataset-collated-20200608232208.csv") 

## Clean data (Using clean_data function)
clean.data <- clean_data(dirty.data)

## Data description
data.description <- describe_data(clean.data)

## Creating dataset with variables selected for this study
final.dataset <- finalize_dataset(clean.data)

## Describing missing data (using describe_missing_data)
descriptions <- describe_missing_data(final.dataset)


##Total patients
total.patients <- nrow(clean.data)

##Missing data
missing.patients <- descriptions[["total.incomplete"]][["n"]]

##Final Data
final.patients <- total.patients - missing.patients

```

```{r datasummarytable, include = TRUE, echo=FALSE, message=FALSE, warning=FALSE}
## Create Sample Characteristics Table
sample.charac <- create_sample_characteristics_table(final.dataset)

## Print table in manuscript
kable(sample.charac$table, caption = sample.charac$caption, format = "pandoc")

```

```{r interaction, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
## EQ-5D Index Score

##Complete patients
hs.complete <- NROW(final.dataset$eq5dhs) - descriptions[["eq5dhs"]][["n"]]

##Mean HS score
hs.mean <- mean(final.dataset$eq5dhs, na.rm = T)

##EQ5D Score by gender
eq5dhs.fem <- mean(final.dataset$eq5dhs[final.dataset$sex=="Female"], na.rm = T)

eq5dhs.mal <- mean(final.dataset$eq5dhs[final.dataset$sex=="Male"], na.rm = T)


##EQ5D Score by Age
eq5dhs.young.fem <- mean(final.dataset$eq5dhs[final.dataset$sex=="Female" & final.dataset$agegrp2=="18-32"], na.rm = T)
eq5dhs.middle.fem <- mean(final.dataset$eq5dhs[final.dataset$sex=="Female" & final.dataset$agegrp2=="33-59"], na.rm = T)
eq5dhs.old.fem <- mean(final.dataset$eq5dhs[final.dataset$sex=="Female" & final.dataset$agegrp2=="60 and above"], na.rm = T)

eq5dhs.young.mal <- mean(final.dataset$eq5dhs[final.dataset$sex=="Male" & final.dataset$agegrp2=="18-32"], na.rm = T)
eq5dhs.middle.mal <- mean(final.dataset$eq5dhs[final.dataset$sex=="Male" & final.dataset$agegrp2=="33-59"], na.rm = T)
eq5dhs.old.mal <- mean(final.dataset$eq5dhs[final.dataset$sex=="Male" & final.dataset$agegrp2=="60 and above"], na.rm = T)

##Creating a model with age and gender
model1 <- lm(eq5dhs ~ age*sex, data = final.dataset)
#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term

summary(model1)
##Interaction not significant 

##Calculating the VIF using the car package
vif(model1)

##It is above 5 for age (7.5) and the interaction factor (8.8)

##Standardizing age and interaction factor by Centering: subtracting by mean and dividing by SD
final.dataset$centred.age <- (final.dataset$age - mean(final.dataset$age, na.rm = TRUE)) / sd(final.dataset$age, na.rm = TRUE)

##New model using centred age
cent.model1 <- lm(eq5dhs ~ centred.age*sex, data = final.dataset)

summary(cent.model1)
##Estimate values do change a bit 

vif(cent.model1)
##But all the VIFs are less than 5

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)


model2 <- lm(eq5dhs ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + centred.age*sex, data = final.dataset)

summary(model2)
##Interaction not significant. Interaction coefficient is -1.13 and after adjusting it is -1.59.

vif(model2)
##All values below 5

##Checking for individual slopes by sex  using the emtrends function

emtrends(model2, ~ sex, var = "centred.age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(model2, pairwise ~ sex, var = "centred.age")

##Interaction of age and gender is not significant. 

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(centred.age = seq(-1.32,3.10, by = 0.44), sex = c("Female","Male")))
plot1 <- emmip(model2, sex ~ centred.age,at = categories1, CIs=TRUE)

##saving plot
jpeg("plot-1.Health-Status.jpeg", quality = 100)
emmip(model2, sex ~ centred.age,at = categories1, CIs=TRUE)
dev.off()

##Overall, with older age it is men who tend to have low QOL, while young women have lower QOL than young men. However, the large overlap of confidence interval indicates EQ5D Health status across age is not statistically different for men and women after adjusting. 

##Dividing age into categories and comparing interaction over sex

##Making age into categorical
final.dataset$agegrp <- cut(final.dataset$age, c(17,32,59,89), labels = c("18-32","33-59","60 and above"))

##Making male the reference group and age group 18-32 the reference group as they had the highest QOL

final.dataset$agegrp2 <- relevel(final.dataset$agegrp, ref = "18-32")
final.dataset$sex2 <- relevel(final.dataset$sex, ref = "Male")

model3 <- lm(eq5dhs ~ sex2*agegrp2, data = final.dataset)

summary(model3)

## Younger men aged (aged 18-25) had significantly higher QOL scores than the middle aged men (aged 33-49). Older women and younger women had better QOL scores than younger women. All other groups had lower scores than younger men.

vif(model3)
##All values are less than 5

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
model4 <- lm(eq5dhs ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset)

summary(model4)

##Calculating CI
model4.ci <- confint.default(model4)
hs <- cbind(coef(model4), model4.ci)


##As before above, younger men aged (aged 18-25) had significantly higher QOL scores than the middle aged men (aged 33-45) (Estimate -4.56). Middle aged women (Estimate 2.27) had much better QOL scores than younger men, while younger women had marginally better QOL scores than younger men (Estimate 0.6). All other groups had lower scores than younger men. 

vif(model4)
##All values are less than 5

##Calculating the values of QOL by all possible combinations of the age and sex categories
cat.values.hs <- emmeans(model4, ~ sex2*agegrp2)

##Estimating the differences in interactions across age and sex categories interaction using contrast function and argument "revpairwise" as Female is the default in the data set

contrast(cat.values.hs, "revpairwise", by ="agegrp2")

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
plot2 <- emmip(model4, sex2 ~ agegrp2, CIs = TRUE)

##Saving plot
jpeg("plot-2.Health-Status.jpeg", quality = 100)
emmip(model4, sex2 ~ agegrp2, CIs = TRUE)
dev.off()

##Middle aged men and Old aged women (33-49) seem to have the lowest QOL Scores.

##EQ5D Domains
## Mobility

##Converting three category outcome to two
final.dataset$eq5dm.new <- recode_factor(final.dataset$eq5dm, "Some Problems" = "Any problem", 
                                "Confined to bed" = "Any problem")


##Creating a model with age and gender using the logistic regression 
m.model1 <- glm(eq5dm.new ~ age*sex, data = final.dataset, family = binomial)

#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term
summary(m.model1)
m.model.ci <- confint.default(m.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(m.model1), m.model.ci))

##For males the odds of having some mobility problems was 1.18 times that of females. For every one year increase in age the odds of some mobility problems increases by 0.98 times (i.e. decreases by 2%).The interaction factor reduces the odds of having some mobility problems by 1% [(1- 0.99*100)].

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
m.model2 <- glm(eq5dm.new ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset, family = binomial)


summary(m.model2)
m.model2.ci <- confint.default(m.model2)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(m.model2), m.model2.ci))

##After adjusting, for males the odds of having some mobility problems was 1.17 times that of females. For every one year increase in age the odds of some mobility problems increases by 0.99 times (i.e. decreases by 1%).The interaction factor reduces the odds of having some mobility problems by 1% [(1- 0.99)*100)].??


##Checking for individual slopes by sex  using the emtrends function
emtrends(m.model2, ~ sex, var = "age")


##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(m.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.0014

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
m.plot1 <- emmip(m.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Mobility.jpeg", quality = 100)
emmip(m.model2, sex ~ age, at = categories1, type = "response", CIs=FALSE)
dev.off()

##Tried the scales package and the scale function in emmip scales = exp_trans(), ylim = c(0, 2) or scales = log10_trans(). It doesn't seem to work

##Using effects package for a stacked plot
jpeg("plot-1.Mobility_Stacked.jpeg", quality = 100)
plot(Effect(focal.predictors = c("age","sex"), mod = m.model2), 
     rug = FALSE,
     style="stacked")
dev.off()

##Using effects package for a split groups plot
jpeg("plot-1.Mobility_Split.jpeg", quality = 100)
plot(Effect(focal.predictors = c("age","sex"), mod = m.model2), 
     rug = FALSE)
dev.off()

##Dividing age into categories and comparing interaction over sex
m.model3 <- glm(eq5dm.new ~ sex2*agegrp2, data = final.dataset, family = binomial)

summary(m.model3)
m.model3.ci <- confint.default(m.model3)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(m.model3), m.model3.ci))

## For younger females (18-32), middle aged females (33-59), older females (60 and above) the odds of having some mobility problems was 0.80, 1.31, and 0.77 times that of younger males (18-32). The odds of middle aged males (33-59) and older males (50 and above) having some mobility problems was 0.70 and 0.54 times than younger males

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
m.model4 <- glm(eq5dm.new ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset, family = binomial)

m.coef <- coef(summary(m.model4))

##p value
##m.p <- pnorm(abs(m.coef[, "t value"]), lower.tail = FALSE) * 2
##m.p.val <- (m.coef <- cbind(m.coef, "p value" = m.p))

##OR
m.model4.ci <- confint.default(m.model4)

##As the estimates are ordered log odds converting them to odds ratios
mobility.odds <- exp(cbind(OR = coef(m.model4), m.model4.ci))

##OR
m.young.fem <- mobility.odds["sex2Female", "OR"]
m.middle.fem <- mobility.odds["sex2Female:agegrp233-59", "OR"]
m.old.fem <- mobility.odds["sex2Female:agegrp260 and above", "OR"]
m.middle.mal <- mobility.odds["agegrp233-59", "OR"]
m.old.mal <- mobility.odds["agegrp260 and above", "OR"]

## After adjusting, for younger females (18-32), middle aged females (33-59), older females (60 and above) the odds of having some mobility problems was 0.78, 1.44, and 0.77 times that of younger males (18-32). The odds of middle aged males (33-59) and older males (60 and above) having some mobility problems was 0.77 and 0.72 times than younger males. All groups had lower odds of having some mobility problems than young males except middle aged women.

##Calculating the values of mobility by all possible combinations of the age and sex categories
 m.cat.values <- emmeans(m.model4, ~ sex2*agegrp2)
 contrast(m.cat.values, "revpairwise", by="agegrp2",adjust="none")

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
m.plot2 <- emmip(m.model4, agegrp2 ~ sex2, type = "response", CIs = TRUE)

##Saving plot
jpeg("plot-2.Mobility.jpeg", quality = 100)
emmip(m.model4, agegrp2 ~ sex2, CIs = FALSE)
dev.off()

##Using effects package for a stacked plot
jpeg("plot-2.Mobility_Stacked.jpeg", quality = 100)
plot(Effect(focal.predictors = c("agegrp2","sex2"), mod = m.model4), 
     rug = FALSE,
     style="stacked")
dev.off()

##Using effects package for a split groups plot
jpeg("plot-2.Mobility_Split.jpeg", quality = 100)
plot(Effect(focal.predictors = c("agegrp2","sex2"), mod = m.model4), 
     rug = FALSE)
dev.off()


##m.plot.new <- emmip(m.cat.values, sex2 ~ agegrp2, CIs=TRUE, plotit=FALSE)
##m.graph <- ggplot(data=m.plot.new, aes(x=agegrp2,y=yvar, fill=sex2)) + geom_bar(stat="identity",position="dodge")


##Self-care

##Converting three category outcome to two
final.dataset$eq5dsc.new <- recode_factor(final.dataset$eq5dsc, "Some Problems" = "Any problem", 
                                "Unable to wash or dress" = "Any problem")

##Creating a model with age and gender
sc.model1 <- glm(eq5dsc.new ~ age*sex, data = final.dataset, family = binomial)

summary(sc.model1)
sc.model.ci <- confint.default(sc.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(sc.model1), sc.model.ci))

##For males the odds of having some self-care problems was 1.59 times that of females. For every one year increase in age the odds of some mobility problems increases by 1%.The interaction factor does not change the odds of having self-care [0.99].

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
sc.model2 <- glm(eq5dsc.new ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset, family = binomial)


summary(sc.model2)
sc.model2.ci <- confint.default(sc.model2)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(sc.model2), sc.model2.ci))

##After adjusting, for males the odds of having some mobility problems was 1.65 times that of females. For every one year increase in age the odds of some mobility problems was increased by 1%. The interaction factor did not change the odds of having self-care problems [0.99].

##Checking for individual slopes by sex  using the emtrends function
emtrends(sc.model2, ~ sex, var = "age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(sc.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.0052

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
sc.plot1 <- emmip(sc.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Self-care.jpeg", quality = 100)
emmip(sc.model2, sex ~ age, at = categories1, type = "response", CIs=FALSE)
dev.off()

##Dividing age into categories and comparing interaction over sex
sc.model3 <- glm(eq5dsc.new ~ sex2*agegrp2, data = final.dataset, family = binomial)

summary(sc.model3)
sc.model3.ci <- confint.default(sc.model3)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(sc.model3), sc.model3.ci))

## For younger females (18-32), middle aged females (33-59), older females (60 and above) the odds of having some problems performing self-care were 0.75, 1.05, and 0.95 times that of younger males (18-32). The odds of middle aged males (33-59) and older males (60 and above) having some problems performing self-care was 0.73 and 0.52 times than younger males. 

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
sc.model4 <- glm(eq5dsc.new ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset, family = binomial)

summary(sc.model4)
sc.coef <- coef(summary(sc.model4))

##p value
##sc.p <- pnorm(abs(sc.coef[, "t value"]), lower.tail = FALSE) * 2
##sc.p.val <- (sc.coef <- cbind(sc.coef, "p value" = sc.p))

##OR

sc.model4.ci <- confint.default(sc.model4)

##As the estimates are ordered log odds converting them to odds ratios
sc.odds <- exp(cbind(OR = coef(sc.model4), sc.model4.ci)) 

##OR
sc.young.fem <- sc.odds["sex2Female", "OR"]
sc.middle.fem <- sc.odds["sex2Female:agegrp233-59", "OR"]
sc.old.fem <- sc.odds["sex2Female:agegrp260 and above", "OR"]
sc.middle.mal <- sc.odds["agegrp233-59", "OR"]
sc.old.mal <- sc.odds["agegrp260 and above", "OR"]

## After adjusting, for younger females (18-32), middle aged females (33-59), older females (60 and above) the odds of having some problems performing self-care was 0.74, 1.15, and 1.05 times that of younger males (18-32). The odds of middle aged males (33-59) and older males (60 and above) some problems performing self-care was 0.82 and 0.66 times than younger males. Middle aged and older females had higher odds of problems with self-care than young males.

##Calculating the values of mobility by all possible combinations of the age and sex categories
 sc.cat.values <- emmeans(sc.model4, ~ sex2*agegrp2, type = "response")

##Estimating the differences in interactions across age and sex categories interaction using pairs function as these are logs of odds

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
sc.plot2 <- emmip(sc.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)

##Saving plot
jpeg("plot-2.Self-care.jpeg", quality = 100)
emmip(sc.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)
dev.off()

##Usual activities

##Converting three category outcome to two
final.dataset$eq5dua.new <- recode_factor(final.dataset$eq5dua, "Some Problems" = "Any problem", 
                                "Unable to perform usual activities" = "Any problem")

##Creating a model with age and gender
ua.model1 <- glm(eq5dua.new ~ age*sex, data = final.dataset, family = binomial)

#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term
summary(ua.model1)
ua.model.ci <- confint.default(ua.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ua.model1), ua.model.ci))

##For males the odds of having some problems in performing usual activities was 1.4 times that of females. For every one year increase in age the odds of some mobility problems increases by 1%.The interaction factor does not affect the odds of having self-care [0.99].

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
ua.model2 <- glm(eq5dua.new ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset, family = binomial)


summary(ua.model2)
ua.model2.ci <- confint.default(ua.model2)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ua.model2), ua.model2.ci))

##After adjusting, for males the odds of having some problems with usual activities was 1.46 times that of females.

##Checking for individual slopes by sex  using the emtrends function
emtrends(ua.model2, ~ sex, var = "age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(ua.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.0053

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
ua.plot1 <- emmip(ua.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Usual-activities.jpeg", quality = 100)
emmip(ua.model2, sex ~ age, at = categories1, type = "response", CIs=FALSE)
dev.off()

##Dividing age into categories and comparing interaction over sex
ua.model3 <- glm(eq5dua.new ~ sex2*agegrp2, data = final.dataset, family = binomial)

summary(ua.model3)
ua.model3.ci <- confint.default(ua.model3)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ua.model3), ua.model3.ci))

## For younger females (18-32), middle aged females (33-59), older females (60 and above) the odds of having some problems performing usual activities were 0.70, 1.31, and 0.86 times that of younger males (18-32). The odds of middle aged males (33-59) and older males (60 and above) having some problems performing self-care was 0.69 and 0.61 times than younger males. 

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
ua.model4 <- glm(eq5dua.new ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset, family = binomial)

summary(ua.model4)
ua.coef <- coef(summary(ua.model4))

##p value
##ua.p <- pnorm(abs(ua.coef[, "t value"]), lower.tail = FALSE) * 2
##ua.p.val <- (ua.coef <- cbind(ua.coef, "p value" = ua.p))


ua.cat.values <- emmeans(ua.model4, ~ sex2*agegrp2)
contrast(ua.cat.values, "revpairwise", by="agegrp2",adjust="none")

##Age and gender effect is not significant for QOL scores in this data set

##OR
ua.model4.ci <- confint.default(ua.model4)

##As the estimates are ordered log odds converting them to odds ratios
ua.odds <- exp(cbind(OR = coef(ua.model4), ua.model4.ci)) 

##OR
ua.young.fem <- ua.odds["sex2Female", "OR"]
ua.middle.fem <- ua.odds["sex2Female:agegrp233-59", "OR"]
ua.old.fem <- ua.odds["sex2Female:agegrp260 and above", "OR"]
ua.middle.mal <- ua.odds["agegrp233-59", "OR"]
ua.old.mal <- ua.odds["agegrp260 and above", "OR"]

## After adjusting, for younger females (18-32), middle aged females (33-59), older females (60 and above) the odds of having some problems performing usual activities was 0.70, 1.45, and 0.90 times than of younger males (18-32). The odds of middle aged males (33-59) and older males (60 and above) some problems performing usual activities was 0.7 and 0.73 times than young males. Only middle aged and older females had lower odds of having problems performing usual activities than young males.

##Calculating the values of mobility by all possible combinations of the age and sex categories
 ua.cat.values <- emmeans(ua.model4, ~ sex2*agegrp2, type = "response")

##Plotting age categorical interaction
ua.plot2 <- emmip(ua.model4, agegrp2 ~ sex2, CIs = FALSE, type = "response")

##Saving plot
jpeg("plot-2.Usual-activities.jpeg", quality = 100)
emmip(ua.model4, sex2~agegrp2, CIs = TRUE, type = "response")
dev.off()

##Pain and discomfort

final.dataset$eq5dpd.new <- recode_factor(final.dataset$eq5dpd, "Moderate Pain" = "Any Pain", 
                                          "Extreme Pain" = "Any Pain")

##Creating a model with age and gender
pd.model1 <- glm(eq5dpd.new ~ age*sex, data = final.dataset, family = binomial)

#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term
summary(pd.model1)
pd.model.ci <- confint.default(pd.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(pd.model1), pd.model.ci))

##For males the odds of having any form of pain or discomfort was 3 times that of females. For every one year increase in age the odds of any pain or discomfort increases by 1%.The interaction factor reduces the odds of  affect the odds of pain or discomfort by 1%.

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
pd.model2 <- glm(eq5dpd.new ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset, family = binomial)

summary(pd.model2)
pd.model2.ci <- confint.default(pd.model2)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(pd.model2), pd.model2.ci))

##After adjusting, for males the odds of having pain or discomfort was 2.9 times that of females. For every one year increase in age the odds of any pain or discomfort increased by 1%.

##Checking for individual slopes by sex  using the emtrends function
emtrends(pd.model2, ~ sex, var = "age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(pd.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.021

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
pd.plot1 <- emmip(pd.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Pain-Discomfort.jpeg", quality = 100)
emmip(pd.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)
dev.off()

##Dividing age into categories and comparing interaction over sex
pd.model3 <- glm(eq5dpd.new ~ sex2*agegrp2, data = final.dataset, family = binomial)

summary(pd.model3)
pd.model3.ci <- confint.default(pd.model3)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(pd.model3), pd.model3.ci))

## For younger females (18-32), middle aged females (33-59), older females (60 and above) the odds of having some pain or discomfort were 0.57, 1.25, and 3.37 times that of younger males (18-32). The odds of middle aged males (33-59) and older males (60 and above) having any pain was 0.57 and 0.54 times than younger males. 

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
pd.model4 <- glm(eq5dpd.new ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset, family = binomial)

summary(pd.model4)

pd.coef <- coef(summary(pd.model4))

##p value
##pd.p <- pnorm(abs(pd.coef[, "t value"]), lower.tail = FALSE) * 2
##pd.p.val <- (pd.coef <- cbind(pd.coef, "p value" = pd.p))

##OR
pd.model4.ci <- confint.default(pd.model4)

##As the estimates are ordered log odds converting them to odds ratios
pd.odds <- exp(cbind(OR = coef(pd.model4), pd.model4.ci)) 

##OR
pd.young.fem <- pd.odds["sex2Female", "OR"]
pd.middle.fem <- pd.odds["sex2Female:agegrp233-59", "OR"]
pd.old.fem <- pd.odds["sex2Female:agegrp260 and above", "OR"]
pd.middle.mal <- pd.odds["agegrp233-59", "OR"]
pd.old.mal <- pd.odds["agegrp260 and above", "OR"]

## After adjusting, for younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having any pain was 1.7, 0.58, and 0.60 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) some problems performing usual activities was 2.1 and 1.9 times than younger males. Only Middle aged and older males had lower odds of having any form of pain or discomfort than younger males.

##Calculating the values of mobility by all possible combinations of the age and sex categories
pd.cat.values <- emmeans(pd.model4, ~ sex2*agegrp2, type = "response")
contrast(pd.cat.values, "revpairwise", by="agegrp2",adjust="none")

##Estimating the differences in interactions across age and sex categories interaction using pairs function as these are logs of odds

##Age and gender effect is not significant for QOL scores in this data set

##Plotting age categorical interaction
pd.plot2 <- emmip(pd.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)

##Saving plot
jpeg("plot-2.Pain-discomfort.jpeg", quality = 100)
emmip(pd.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)
dev.off()

##Anxiety & Depression

final.dataset$eq5dad.new <- recode_factor(final.dataset$eq5dad, "Moderately Anxious/depressed" = "Any Anxiety/Depression", "Extremely Anxious/depressed" = "Any Anxiety/Depression")

##Creating a model with age and gender using the polr function
ad.model1 <- glm(eq5dad.new ~ age*sex, data = final.dataset, family = binomial)

#Interaction term indicated by a * is equivalent to adding the lower order terms to the interaction term
summary(ad.model1)
ad.model.ci <- confint.default(ad.model1)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ad.model1), ad.model.ci))

##For males the odds of having any form of anxiety and depression  was 1.49 times that of females. For every one year increase in age the odds of any anxiety and depression increases by 1%.The interaction factor does not seem to affect odds of anxiety or depression [0.99].

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
ad.model2 <- glm(eq5dad.new ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + age*sex, data = final.dataset, family = binomial)

summary(ad.model2)
ad.model2.ci <- confint.default(ad.model2)

##As the estimates are ordered log odds converting them to odds ratios
exp(cbind(OR = coef(ad.model2), ad.model2.ci))

##After adjusting, for males the odds of having any anxiety or depression was 1.44 times that of females. 

##Checking for individual slopes by sex  using the emtrends function
emtrends(ad.model2, ~ sex, var = "age")

##The slopes are not significant for males or females

##Testing difference in slope of male and slope of females using the pairwise argument
emtrends(ad.model2, pairwise ~ sex, var = "age")

##Interaction of age and gender is not significant. The adjusted interaction factor is 0.0048

##Plotting the interaction using the emmip function

##Defining the categories 
(categories1 <- list(age = seq(17,89, by = 5), sex = c("Female","Male")))
ad.plot1 <- emmip(ad.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)

##saving plot
jpeg("plot-1.Anxiety-Depresssion.jpeg", quality = 100)
emmip(ad.model2, sex ~ age, at = categories1, type = "response", CIs=TRUE)
dev.off()

##Dividing age into categories and comparing interaction over sex
ad.model3 <- glm(eq5dad.new ~ sex2*agegrp2, data = final.dataset, family = binomial)

summary(ad.model3)

ad.coef <- coef(summary(ad.model3))

##p value
#ad.p <- pnorm(abs(ad.coef[, "t value"]), lower.tail = FALSE) * 2
#ad.p.val <- (ad.coef <- cbind(ad.coef, "p value" = ad.p))

##OR
ad.model3.ci <- confint.default(ad.model3)

##As the estimates are ordered log odds converting them to odds ratios
ad.odds <-exp(cbind(OR = coef(ad.model3), ad.model3.ci))

## For younger females (18-32), middle aged females (33-59), older females (60 and above) the odds of having any anxiety and depression were 0.62, 1.8, and 1.21 times that of younger males (18-32). The odds of middle aged males (33-59) and older males (60 and above) having any anxiety and depression was 0.66 and 0.60 times than younger males. 

#Adjusting for Mechanism of injury, Mode of transport, Transfer status, type of injury, GCS, Respirator Rate, systolic blood pressure, heart rate, oxygen saturation)
ad.model4 <- glm(eq5dad ~ moi.grouped + mot.grouped + tran + tyi + gcs + rr + hr + sbp + hr + spo2 + sex2*agegrp2, data = final.dataset, family = binomial)

summary(ad.model4)
ad.model4.ci <- confint.default(ad.model4)

ad.cat.values <- emmeans(ad.model4, ~ sex2*agegrp2)
contrast(ad.cat.values, "revpairwise", by="agegrp2",adjust="none")

##Age and gender effect is not significant for QOL scores in this data set

##As the estimates are ordered log odds converting them to odds ratios
ad.odds <- exp(cbind(OR = coef(ad.model4), ad.model4.ci)) 


##OR
ad.young.fem <- ad.odds["sex2Female", "OR"]
ad.middle.fem <- ad.odds["sex2Female:agegrp233-59", "OR"]
ad.old.fem <- ad.odds["sex2Female:agegrp260 and above", "OR"]
ad.middle.mal <- ad.odds["agegrp233-59", "OR"]
ad.old.mal <- ad.odds["agegrp260 and above", "OR"]


## After adjusting, for younger females (18-32), middle aged females (33-49), older females (50 and above) the odds of having any anxiety and depression was 1.39, 0.48, and 1.10 times that of younger males (18-32). The odds of middle aged males (33-49) and older males (50 and above) some anxiety and depression was 1.59 and 0.84 times than younger males. Only Older males and middle aged females had lower  odds of having any form of anxiety and depression than younger males.

##Calculating the values of mobility by all possible combinations of the age and sex categories
ad.cat.values <- emmeans(ad.model4, ~ sex2*agegrp2, type = "response")

##Plotting age categorical interaction
ad.plot2 <- emmip(ad.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)

##Saving plot
jpeg("plot-2.Anxiety-Depression.jpeg", quality = 100)
emmip(ad.model4, sex2 ~ agegrp2, type = "response", CIs = TRUE)
dev.off()

```

Out of a total of `r total.patients` trauma patients enrolled, `r missing.patients` were excluded for missing data. The final cohort of `r final.patients` included in this study had a median age of `r data.description$age.median` years (IQR: `r data.description$age.iqr.upper`-`r data.description$age.iqr.lower`) with `r data.description$proportion.male` per cent of patients were male. The most common mechanisms of injury were  `r data.description$moi1` (`r data.description$p.road` per cent). Majority of the injuries were blunt (`r data.description$proportion.blunt` per cent) and `r data.description$proportion.transferred` per cent were transferred to the study hospitals. The all cause 30-day mortality was `r  data.description$p.outcome.dead` per cent. The median severity score was x. Details of the study population is given in Table 2.


EQ-5D Scores
------------
`r hs.complete` completed the EQ-5D questionnaire at 3 months after discharge. The mean EQ5D index score was 76 (SD = 20.5). Three months after discharge, young females (18-32 years) had the highest score while younger males reported the lowest score. Just over half the patients (54.8%) reported no problems with mobility while two-thirds reported no problems with self-care. While less than half the patients (43.0%) could carry on usual activities without any problems and only one-third of the patients (34.7%) reported no pain or discomfort after three-months of post-discharge. Around 40% of the patients reported experiencing some form of anxiety or depression. Again, the proportion of young males reporting any problems across all the five domains was the lowest while the proportion  middle-aged females reported experiencing the problems across all the five domains was the highest. The overall EQ-5D scores are provided in Table 3.

EQ-5D in relation to gender and age
-------------------------------
Overall females had a slightly higher mean EQ5D index score (`r eq5dhs.fem`) than males (`r eq5dhs.mal`). On running a linear regression of index scores across sex across age after adjusting for injury etiology, vitals, and severity, young males aged (aged 18-32) performed better than middle-aged (aged 33-59)  and old  (aged 60 and above) males. On the other hand, old females had higher scores than young and middle aged females. Middle-age group had the lowest performance, with middle-aged faring the worst (Fig 1).

Regression analysis in the mobility dimension keeping young males as the reference group, shows that the adjusted odds of reporting mobility problems was `r m.young.fem` in young females (aged 18-32), `r m.middle.fem` middle-aged females (aged 33-59), and `r m.old.fem` old females (aged 60 and above). In middle-aged males (aged 33-49) the odds were `r m.middle.mal` and in old males (aged 50 and above) the odds were `r m.old.mal`. Thus, all groups had higher odds of reporting some mobility problems than young males except middle-aged females. 

Assessing the regression analysis in the self-care dimension, showed the adjusted odds of reporting problems performing self-care tasks was `r sc.young.fem` in young females, `r sc.middle.fem` in middle-aged females, and `r sc.old.fem` old females compared to young males. On the other hand, the odds of reporting these problems in middle-aged males was `r sc.middle.mal` and in old males it was `r sc.old.mal`times than young males. Only middle aged and older females had higher odds of reporting problems with self-care than young males.

In the dimension of being able to perform usual activities, regression analysis indicated that the adjusted odds of reporting any problems in performing was `r ua.young.fem` in young females, `r ua.middle.fem` in middle aged females, and `r ua.old.fem` old females with respect to young males. The odds of reporting these problems were `r ua.middle.mal` in middle aged males and  `r ua.old.mal` times in in old males. Thus, middle aged females had higher odds of having problems performing usual activities than young males.

The adjusted odds of reporting any form of pain and discomfort were `r pd.young.fem` in young females, `r pd.middle.fem` in middle-aged females, and `r pd.old.fem` old females compared to young males. Among middle aged and old males these odds were `r pd.middle.mal` and  `r pd.old.mal` respectively. Thus, middle aged and old females had higher odds of reporting pain than young males. 

In the anxiety and depression dimension, keeping young males as the reference group, the adjusted odds of reporting any form of anxiety or depression were `r ad.young.fem` in young females, `r ad.middle.fem` middle-aged females, and `r ad.old.fem` old females. In middle-aged males the odds was `r ad.middle.mal` and in old males the odds was `r ad.old.mal`. So, younger females and older males had higher  odds of reporting having any form of anxiety and depression than young males. 

Though the odds of reporting any problems across the EQ5D domains were not statistically significant, middle aged females had the higher odds of reporting problems with across all domains except anxiety and depression (Table 4). The interaction of age and gender on the index EQ5D score and across its 5 domains was found to be not statistically significant.  

Discussion
==========

<!-- Need to rework this section based on the new results.-->
This study examined the interaction of age and gender on QOL in trauma patients discharged from four tertiary-care hospital in urban India at 3-months after discharge. We did not find statistically significant differences across all age groups and gender as well as no interaction effect between age and gender in this population. The sub-group analysis showed differences in QOL between males and females across young (18-32 years), middle-aged (33-59), and older (60 years and above) groups. In each of the five EQ5D domains, odds of reporting problems varied between the groups. We found middle-aged men report significantly lower QOL scores and was due to having higher odds of reporting problems in mobility, anxiety or depression, and especially pain, and usual activities. 

The mean EQ5D index score of 76 of trauma patients in this study is lower than the general population norm for South Asia of 85 [@Kularatna2014]. A recent study of trauma patients from in north India, reported index scores of around 90 at 3-months post-injury, lower than our findings [@Jagnoor2020]. This could be because two of the three study sites of that study were secondary-care hospitals, which treat less severe trauma patients. The score is however, higher than other studies which report scores between  48 to 66 for trauma patients at 3-months [Spronk2019; @Wihlke2021]. This indicates that, even after 3-months since discharge, trauma patients still face limitations to achieve full recovery for a long-time after trauma. Pain and usual activities were the key domains of poor QOL in this study, with nearly two-thirds of participants reporting some form of pain at 3-months. This is consistent with other literature on similar cohorts [@Jagnoor2020; Spronk2019; @Wihlke2021]. 

In this study women reported slightly better overall QOL than men, which is consistent with the study from North India [@Jagnoor2020]. Evidence on this remains mixed with some studies reporting lower scores for women [@Rissanen2020] while other report lower scores for men [@Jagnoor2020]. Studies that report women outperforming men also tend to have a smaller proportion of women in their study sample. In this study, the middle-aged cohort, seem to be having the odds for the poorest QOL outcomes. The middle-aged men had the lowest EQ-5D index scores than the other groups, followed by middle-aged women (coefficients respectively of −4.86 (95% CI −9.07; −0.65) and −2.01 (95% CI -11.41; 7.37)). Better QOL was reported after 3-months following the injury in younger age adults as well as those over 60 years. This is in contrast with findings from high-income settings where older adults perform worse [@Gabbe2017; @Wihlke2021] but similar to the findings from India [@Jagnoor2020]. Better scores among the younger age groups older could be due to physiological factors and among the old it could perhaps be associated with the cultural context of care by informal caregivers leading to overall better health [@Ghosh2020]. 

Interestingly, our study shows that while middle-aged and older men had the highest odds of reporting problems in the domains of anxiety, pain and usual activities than other groups. Middle-aged and older women had the lowest odds of reporting any problems across the domains. This could be because of the gendered norms around health-seeking behaviour that could affect reporting of health problems as well [Agrawal2014; @NationalSampleSurveyOffice2019; @Das2018]. This cohort also had the lowest odds of reporting having problems in performing usual activities. This could be due to more men and younger women being part of the the formal workforce, which is deemed in higher value than domestic chores, that older women engage in [@UnitedNationsPopulationFund2017]. These findings are in contrast to other studies from high-income settings show that women report more problems in the pain, anxiety, and usual activities domain than men [@Gabbe2017].It was only in the domain of mobility that young men had the highest odds of reporting problems than other groups. This could be due to the severity of the injuries as well as this cohort engaging in movement and travel compared to other groups.

The findings from our study indicate that men and women do not recover QOL at 3-months post-discharge the same way across different age groups after trauma. The reasons for these differences in different Indian contexts should be explored in detail in future studies. Additionally, future research should focus on specific cohorts of trauma such as traumatic brain injury or spinal cord injuries and road traffic injuries or falls to study the differences in QOL outcomes and the reasons for the same. 

Limitations
The EQ5D tool has not been validated for trauma patients in the Indian context and conducting a telephonic could have affected the ability of participants to understand the questions and thereby the reported responses. However, the tool has been validated in non-trauma settings and culturally similar populations in South Asia. Additionally, the tool has details on how to collect the information over the telephone. This study only captures those who responded during the follow-up calls, leaving out drop-outs. Another limitation is using cross-sectional data that does not allow in-depth  analysis of causal relationships of the associations detected in this study. Longitudinal studies in the future would helping further determining the causal association between age and gender with QOL in low resource settings.


Conclusion
==========



References
==========
